---
title: "The Ten Biggest Storms and their Impact in the Health and Economy of the Population: 1950:2011"
author: "Luis Urbina"
date: "Wednesday, May 20, 2015"
output: html_document
---
## I    SYNOPSIS
.

Storm Data is an official publication of the NOAA, and it documents the
occurrence of sifnificant weather phenomena causing loss of life, injuries,
and significant property damage. Based on the Storm Data database covering
the years from 1950 to 2011, the purpose of this report is to determine
the severe weather events that cause the biggest harm to the population
health. It also is determine the events that have the biggest impact in 
the economic aspect.   

.

## II   DATA PROCESSING 
.

### 2.1) Loading and preprocessing the data
The given the database file 'repdata-data-StormData.csv.bz2', is loaded 
and the dataset is named **data**. In order to sanitize the database, 
blank cells are replaced with the NA value. 

```{r, echo=TRUE, cache=TRUE}
    data = read.csv(bzfile('repdata-data-StormData.csv.bz2'),
                    header=T, na.strings=c("", "NA"))
```
.
It is also necessary to eliminate duplicate rows if they exist.

```{r, echo=TRUE, cache=TRUE}
    data <- data[,-37]
    data <- unique(data)    
```
.

### 2.2) Exploring the dataset behavior.

```{r, echo=TRUE, cache=TRUE}
    str(data)
```

Above is shown the big picture of the dataset. It has 902,297 observations
and 37 variables. Storm events information is given in details about the 
type of event, tornado, flood, etc,; location of the event occurrency, 
state and county names and lattitude and longitude; date and time of 
occurrency; impact in the population health and economic cost of the 
phenomenum. 

Official documents from NOAA indicates that the the variables **FATALITIES** 
and **INJURIES** are directly related with the populationn health harm 
caused for the storm events. They also indicate that the variables 
**PROPDMG**, **PROPDMGEXP**, **CROPDMG**, and **CROPDMGEXP** are directly
related with the economic impact of the phenomenos ocurrency. 

So, the research project is mainly oriented to find out the relation of those
variables with the storm event type, the location and date of ocurrency. 
other minor details are not taken in consideration for the purpose of the 
research.

.

### 2.3) Exploring the variables of interest: people health harm
.

```{r, echo=TRUE, cache=TRUE}
    summary(data[,c(23,24)])
```
.
Above is shown a summary of the main statistic characteristics of the 
**FATALITIES** and **INJURIES** variable. For the nature of the storm 
events, natural phenomenums, the nature of the distribution of these 
variables may be considered discrete. 

With a median equal to zero and a mean very far away of the maximum 
value of the serie, for both variables, a lecture over the statistic summary
indicates that both variables have a kind of chi distribution behavior. 

Two ratios about the incidence of the biggest event over the variable
population, the **fatalities ratio** and the **injuries ratio**, are 
calculated. These are indicators of the impact of the maximum number 
of fatalities and injuries caused for the biggest event over the total 
of fatalities and injuries caused for the total number of events. 

```{r, echo=TRUE, cache=TRUE}
    max(data$FATALITIES) * 100 / sum(data$FATALITIES)
```

```{r, echo=TRUE, cache=TRUE}
    max(data$INJURIES) * 100 / sum(data$INJURIES)
```

Considering tha the storm event database has registred 902,297 events,
the mean of the variables, (0.0168 and 0.1557 for fatalities and injuries
respectively), and the fatalities and injuries ratio, calculated above, 
also indicates a large number of events of low impact and few events
of big impacts.

So this research will conduct to determine the events with the biggest
people health harm.

.


### 2.4) Histograma: Fatalities and Injuries variables.


The **histogram of fatalaties** in an event.
```{r, echo=TRUE, cache=TRUE, fig.align='center'}
    old.par <- par(mfrow=c(1, 2))
    hist(unique(data$FATALITIES), breaks=25, col='red',
            main="People Health Harm",
            xlab = 'Died People')
    hist(unique(data$INJURIES), breaks=25, col='green',
            main="People Health Harm",
            xlab = 'Injuried People')
    par(old.par)
```

.

### 2.5) Exploring the variables of interest: economic impact
.

```{r, echo=TRUE, cache=TRUE}
    summary(data[,c(25,27)])
```

A summary of the main statistic characteristics of the **PROPDMG** and the
**CROPDMG** variables is shown above. The mean of the variables PROPDMG 
and CROPDMG, 12.06 and 1.527 respectively, the median with value equal to 
zero, and having the max value very far away (5000, 990 respectively), 
may indidates that the values of these variables may be following a 
chi square distribution. 

It means that there are a lot of events with very low economic impact 
and a few storm events with high economic impact. Ratios about the cost 
impact of the biggest event over the total cost is calculated in the 
**property ratio** and the **crop ratio**. These are indicators of 
the impact of the maximum cost in property and crop damaged caused for 
the biggest event over the total of damage cost caused for 
the total number of events.

```{r, echo=TRUE, cache=TRUE}
    max(data$PROPDMG) * 100 / sum(data$PROPDMG)
```

```{r, echo=TRUE, cache=TRUE}
    max(data$CROPDMG) * 100 / sum(data$CROPDMG)
```

Considering tha the storm event database has registred 902,297 events,
the mean of the variables, (0.0168 and 0.1557 for fatalities and injuries
respectively), and the fatalities and injuries ratio, calculated above, 
also indicates a large number of events of low impact and few events
of big impacts.

.

```{r, echo=TRUE, cache=TRUE}
    summary(data[,c(26,28)])
```

The **PROPDMGEXP** and **CROPDMGEXP** variables are multiplier factors 
helping to quantify the property and crop damage. Conversely to the other
variables discussed above, these variables are of the factor type. They
also have NA values and will requiere a specific treatment for using in
the calculation process. Notice that the letter **B** means a billion 
multiplier, for example. The **NA** values will be replaced with a value
equal to **1** in order to allow the quantification of the total cost of 
the damatge in both property and crop. 

So this research will conduct to determine the events with the biggest
people health harm.  
















.

### 2.6) The Ten Biggest Storms
.

The impact of the storms in the health of the population is determinated 
by selecting those with the high number of fatalities and injuries. It has
been filtered from the database all the storms that have registered 500 
or more number of fatalities or injuries. On the other hand, to determine 
the biggest storms with economic impact, from the database has been selected 
the storms that have as multiplier the factor "B" wich belong to the level "billion".

.

```{r, echo=TRUE, cache=TRUE}
tidy <- data[,c(2,5,7,8,23,24,25,26,27,28)]

dmg <- rbind(tidy[tidy$FATALITIES > 500 ,], 
                tidy[tidy$INJURIES > 500 ,],
                tidy[tidy$PROPDMGEXP %in% "B",],
                tidy[tidy$CROPDMGEXP  %in% "B",])
#summary(dmg[,c(4,5,6,7,8,9)])
dim(dmg)
```
.
With this criteria is created a tidy dataset which include only the 
variable of interest. This criteria selection give to us a sub-data set
with 66 registers including the storms and weather phenomenas that have 
caused more than 500 injured people, or more than 500 died people, or
have caused damage in properties, or damage in crops in the order of 
billions of dollar.

.

### 2.7) Quantifying the total damage
.
In order to quantified the impact of the above selected storms in the 
health and economy of the population, two variables are been created. The
new **TOT.HARM** variable quantified the total number of persons that have
suffer injured or fatalities caused by a storm. It adds the values of the
FATALITIES and INJURIES variables. 


```{r, echo=TRUE, cache=TRUE}
t <- dmg
t <- apply(t, 1, function(x){ifelse(x=="K", 1000, x)})
t <- apply(t, 1, function(x){ifelse(x=="M", 1000000, x)})
t <- apply(t, 1, function(x){ifelse(x=="B", 1000000000, x)})
t[is.na(t)] <- 1
t <- t(t)
t <- as.data.frame(t)

t$tot.harm <- as.numeric(levels(t$FATALITIES)[t$FATALITIES]) +
                as.numeric(levels(t$INJURIES)[t$INJURIES])

t$tot.dmg <- as.numeric(levels(t$PROPDMG)[t$PROPDMG]) * 
            as.numeric(levels(t$PROPDMGEXP)[t$PROPDMGEXP]) +
            as.numeric(levels(t$CROPDMG)[t$CROPDMG]) * 
            as.numeric(levels(t$CROPDMGEXP)[t$CROPDMGEXP])

summary(t[,c(11,12)])
```

```{r, echo=TRUE, cache=TRUE}
top.harm <- t[order(-t$tot.harm),]
head(top.harm[,c(1,2,3,4,12,11)], n=10)
```

```{r, echo=TRUE, cache=TRUE}
top.dmg <- t[order(-t$tot.dmg),]
head(top.dmg[,c(1,2,3,4,11,12)], n=10)
```


.



## III  RESULTS
#### 1)
#### 2)
#### 3)
#### 1.3) Imputing missing values
