---
title: "Peer Assessment 02"
author: "Luis Urbina"
date: "Wednesday, May 20, 2015"
output: html_document
---

## Loading and preprocessing the data
#### 1.1) Loading the data
Given the data file 'repdata-data-StormData.csv.bz2', the data is 
loaded using the `read.csv()` function. The dataset is named **data**.

```{r, echo=TRUE, cache=TRUE}
    data = read.csv(bzfile('repdata-data-StormData.csv.bz2'))    
```

#### 1.2) Exploring the dataset behavior.

```{r, echo=TRUE, cache=TRUE}
    dim(data)
    str(data)
    summary(data)
```

#### 1.3) Imputing missing values

The **Total Number of Missing Values**
```{r, echo=TRUE, cache=TRUE}
    tidy.NA <- data[,c(1,2,8,22,23)]
    na.identifier <- apply(tidy.NA, 1, function(x){any(is.na(x))})
    total.NA <- sum(na.identifier)
    total.NA
```


## Population health harm and greatest economic consecuences
The NA values are going to be ignnored in this part of the analysis.

#### 2.1) Total population health harm - MAG
 
The **MAG maximo** in an event.
```{r, echo=TRUE, cache=TRUE}
    max(data$MAG)
```


The **media of the MAG** acrross the time.
```{r, echo=TRUE, cache=TRUE}
    mean(data$MAG)
```


The **median of the MAG** acrross the time.
```{r, echo=TRUE, cache=TRUE}
    median(data$MAG)
```


The **histogram of MAG** acrross the time.
```{r, echo=TRUE, cache=TRUE}
    hist(unique(data$MAG), breaks=25, col='red')
```



#### 2.2) Total population health harm - FATALITIES
 
The **total number of fatalaties** suffered for people.
```{r,echo=TRUE}
    sum(data$FATALITIES)
```


The **maximo number of fatalaties** in an event.
```{r,echo=TRUE}
    max(data$FATALITIES)
```


The **media of the FATALITIES** acrross the time.
```{r,echo=TRUE}
    mean(data$FATALITIES)
```


The **median of the FATALITIES** acrross the time.
```{r,echo=TRUE}
    median(data$FATALITIES)
```


The **histogram of fatalaties** in an event.

```{r,echo=TRUE}
    hist(unique(data$FATALITIES), breaks=25, col='red')
```

#### 2.3) Total population health harm - INJURIES
 
The **total number of INJURIES** suffered for people.
```{r,echo=TRUE}
    sum(data$INJURIES)
```


The **maximo number of INJURIES** in an event.
```{r,echo=TRUE}
    max(data$INJURIES)
```


The **media of the INJURIES** acrross the time.
```{r,echo=TRUE}
    mean(data$INJURIES)
```


The **median of the INJURIES** acrross the time.
```{r,echo=TRUE}
    median(data$INJURIES)
```


The **histogram of INJURIES** in an event.

```{r,echo=TRUE}
    hist(unique(data$INJURIES), breaks=25, col='green')
```
