---
title: "Peer Assessment 02"
author: "Luis Urbina"
date: "Wednesday, May 20, 2015"
output: html_document
---
## I    SYNOPSIS
.

## II   DATA PROCESSING 
.

### Loading and exploring the data
.

#### 2.1) Loading the data
Given the data file 'repdata-data-StormData.csv.bz2', the data is 
loaded using the `read.csv()` function. The dataset is named **data**.

```{r, echo=TRUE, cache=TRUE}
    data = read.csv(bzfile('repdata-data-StormData.csv.bz2'))    
```
.

#### 2.2) Exploring the dataset behavior.

```{r, echo=TRUE, cache=TRUE}
    str(data)
```

The `str()` function give as a big picture of the dataset such as the 
variables name, total number of observation (902297 observations), and 
the total number of variables (37). It also indicates the varible type
of each variable and visualize that the **REFNUM** is a unique value 
variable type and may be used as index.

.

### Events with Population Health Harm 
.

#### 2.3) Sanitizing the working variables
According to the document, the variables **FATALITIES** and **INJURIES** 
are directly related with the populationn health harm caused for the storm
events. So, before of make a variable analysis, it is performed a sanitizing
of the working variables involved. 


Identifying **blank values**, **NA**, and **NULL** in the working sub-dataframe.
```{r, echo=TRUE, cache=TRUE}
    tidy.fat.inj <- data[,c(1,2,8,22,23)]
    blank.id <- apply(tidy.fat.inj, 1, function(x){ifelse(x=="", TRUE, FALSE)})
    total.blank <- sum(blank.id)
    total.blank
```

```{r, echo=TRUE, cache=TRUE}
    na.id <- apply(tidy.fat.inj, 1, function(x){{any(is.na(x))}})
    total.na <- sum(na.id)
    total.na
```

```{r, echo=TRUE, cache=TRUE}
    null.id <- apply(tidy.fat.inj, 1, function(x){{any(is.null(x))}})
    total.null <- sum(null.id)
    total.null
```

.

#### 2.4) Analisis of variables: Fatalities and Injuries.

A brief analysis of these variables is next.

```{r, echo=TRUE, cache=TRUE}
    summary(data[,c(23,24)])
```

For the nature of the events, natural phenomenum, the nature of the 
distribution of these variables may be considered discrete. With a 
median equal to zero and a mean very far away of the maximum 
value of the serie, for both variables, a lecture over the summary 
indicates that both variables have a kind of chi distribution behavior. 


The **fatalities ratio** and the **injuries ratio** are indicators of 
the impact of the maximum number of fatalities and injuries caused for 
the biggest event over the total of fatalities and injuries caused for 
the total number of events. 

```{r, echo=TRUE, cache=TRUE}
    max(data$FATALITIES) * 100 / sum(data$FATALITIES)
```

```{r, echo=TRUE, cache=TRUE}
    max(data$INJURIES) * 100 / sum(data$INJURIES)
```

Considering tha the storm event database has registred 902,297 events,
the mean of the variables, (0.0168 and 0.1557 for fatalities and injuries
respectively), and the fatalities and injuries ratio, calculated above, 
also indicates a large number of events of low impact and few events
of big impacts.

So this research will conduct to determine the events with the biggest
people health harm.  

The **histogram of fatalaties** in an event.
```{r, echo=TRUE, cache=TRUE}
    hist(unique(data$FATALITIES), breaks=25, col='red')
```

The **histogram of INJURIES** in an event.

```{r, echo=TRUE, cache=TRUE}
    hist(unique(data$INJURIES), breaks=25, col='green')
```

.

#### 2.5) Identification of events with big impact 

```{r, echo=TRUE, cache=TRUE}
    data.fat.inj <- rbind(data[(data$FATALITIES > 500),], 
                          data[(data$INJURIES > 500),])
    data.fat.inj$TOTAL <- (data.fat.inj$FATALITIES) + 
                            (data.fat.inj$INJURIES)
    top.fat.inj <- data.fat.inj[order(-data.fat.inj$TOTAL),]
    
    head(top.fat.inj[,c(2,7,8,23,24,38)], n=8, row.names=FALSE)
```

.

#### Events with Greatest Economic Consecuences

According to the document, the variables **PROPDMG**, **PROPDMGEXP**, 
**CROPDMG**, and **CROPDMGEXP** are directly related with the economic damage caused by the storm events. A brief analysis of these variables is next.

Identifying **blank values**, **NA**, and **NULL** in the working sub-dataframe.
```{r, echo=TRUE, cache=TRUE}
    tidy.pro.dam <- data[,c(25,27)]
    blank.id <- apply(tidy.pro.dam, 1, function(x){ifelse(x=="", TRUE, FALSE)})
    total.blank <- sum(blank.id)
    total.blank
```

```{r, echo=TRUE, cache=TRUE}
    tidy.pro.dam <- data[,c(26,28)]
    blank.id <- apply(tidy.pro.dam, 1, function(x){ifelse(x=="", TRUE, FALSE)})
    total.blank <- sum(blank.id)
    total.blank
```


```{r, echo=TRUE, cache=TRUE}
    na.id <- apply(tidy.pro.dam, 1, function(x){{any(is.na(x))}})
    total.na <- sum(na.id)
    total.na
```

```{r, echo=TRUE, cache=TRUE}
    null.id <- apply(tidy.pro.dam, 1, function(x){{any(is.null(x))}})
    total.null <- sum(null.id)
    total.null
```

The property and crop damage columns do not have blank, NA, or NULL values. 
The property and crop damage extension columns also do not have NA or NULL 
values, but they have blank values. 

.

#### 2.3) Analisis of variables
.

A brief analysis of the variables related to the econmic damage is next.
.

```{r, echo=TRUE, cache=TRUE}
    summary(data[,c(26,28)])
```

The PROPDMGEXP and CROPDMGEXP variables are multiplier factors to quantify 
the property and crop damage variables. Recalling the sanitizing process made 
above, these variables have have blank values. So, the blank values are going 
to be replacing with the value equal to **1** in order to quantify the total 
value of damage in both property and crop. Next is a the process to filling 
up the blank value per "1".
.



```{r, echo=TRUE, cache=TRUE}
    summary(data[,c(25,27)])
```

The mean of the variables PROPDMG and CROPDMG, 12.06 and 1.527 respectively, 
the median with value equal to zero, and having the max value very far away 
(5000, 990 respectively), indidates that the values of these variables may 
be following a chi square distribution. It means that there are a lot of events
with very low economic impact and a few storm events with a high economic 
impact. 
.



The **property ratio** and the **crop ratio** are indicators of 
the impact of the maximum cost in property and crop damaged caused for 
the biggest event over the total of damage cost caused for 
the total number of events.

```{r, echo=TRUE, cache=TRUE}
    max(data$FATALITIES) * 100 / sum(data$FATALITIES)
```

```{r, echo=TRUE, cache=TRUE}
    max(data$INJURIES) * 100 / sum(data$INJURIES)
```

Considering tha the storm event database has registred 902,297 events,
the mean of the variables, (0.0168 and 0.1557 for fatalities and injuries
respectively), and the fatalities and injuries ratio, calculated above, 
also indicates a large number of events of low impact and few events
of big impacts.

So this research will conduct to determine the events with the biggest
people health harm.  










## III  RESULTS
#### 1)
#### 2)
#### 3)
#### 1.3) Imputing missing values


